---
title: "Project Management and Other Topics"
author: "Sky Onosson, Coding Discussion Group coordinator"
date: "May, 2020"
output: 
  html_document: 
    theme: yeti
    highlight: pygments
    number_sections: yes
    toc: yes
---

```{r setup, include=F}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

# Naming

In dealing with projects on your computer, you will have to provide names to a variety of items:

* folders
* files
* variables
* etc.

For projects which can have a long shelf-life, i.e. that you will be working with over weeks or months while also dealing with other projects, it is essential to develop good naming practices. While a short, abbreviated name might make sense and be perfectly clear in the moment, when you come back to it after some time and/or after working on something else, you might not recall why you chose that particular name or what it means.

When it comes to choosing a good name for a digital object, I feel that there are two large principles or constraints at work, both of which are important but which may come into conflict with each other. These are **brevity** and **clarity**. Name selection usually involves finding a good balance between these.

## Brevity

Simply put, **brevity** involves choosing short names, which can minimally be as short as one character such as `a` or `x`. I would include most abbreviations in the category of short names, as well. 

Short names have a number of advantages:

* faster to type
* easy to select and remember in the moment
* stand out from text composed of longer strings

When creating project files, code variables, etc. it is often convenient to use short names at first, for the above reasons. For short-term projects, the use of short names may not cause any difficulties, and the advantages listed above are good reasons to use them in the right contexts.
But in longer timeframes, short names can cause a lot of problems due to their opaque quality.

## Clarity

**Clarity** in naming means that names are clear, which I attribute to (at least) two factors:

1. they are easily **decomposable**
1. they are **meaningful**

**Decomposability** means that you can easily recognize what the parts of a name are. Abbreviations fall under "short names" because they aren't easy to decompose -- you might know the full form of S.C.U.B.A. but its not very decomposable as the word *scuba*.

**Meaningfulness** is whether you can get meaning out of a name. Abbreviations are not very meaningful in their abbreviated form, but neither are names such as "CoursePaperFinalDraft". While there is some meaning, a file with this name is not easily distinguished from other similar files.

Names with good **clarity** will utilize both of these elements. A name such as "University_of_Manitoba_Masters_CourseName_Final_Term_Paper_v3" is both easily decomposable and highly meaningful. However, it obviously runs counter to the principle of **brevity**.

## Compromise in naming

The best names strike a middle ground, in that they are both brief and clear. What this means in practice depends entirely on context. For folders which contain projects, I find it most useful to use the hierarchical structure of the filesystem to build in decomposability and meaning, while keeping the folder and file names more brief. Subfolders are your friend!

```{r, eval=F}
Thesis/

  Data/

    Original_Files/
      File_A.wav
      File_B.wav

    Annotated_Files/
      File_A_annotated.textgrid
      File_B_annotated.textgrid
  
  R_Scripts/
    Stats.R

  Documents/
    Thesis_v1
    Thesis_v2

===============================

Course_A/

  Notes/
    Sep_10.doc
    Sep_12.doc

  Papers/
    Term_paper1_v1.doc

===============================

Course_B/

etc...
```

When it comes to variable names in code, you are not working within a filesystem, so your options are different.
One advantage with a lot of editors like RStudio is their autocomplete functionality, so that you very often don't need to fully type out longer names.
As a personal preference, I usually deal with variable names in two categories:

1. Names that will be only used in a restricted part of my code, and which are fairly easy to interpret within that context. For example, I might be running a for-loop in a section of code which takes a series of files, one after the other, and performs some operation, such as reading them into memory. In these cases, I will very often favour **brevity** over **clarity**, and might use a one-letter variable name. Often within a for-loop, the loop variable will be used to select items from a vector or list which you have already established, going through the list one item after another as the sequence progresses.
```{r, eval=F}
file_list <- c("file1.csv", "file2.csv")

for (f in 1:2) {read.csv(file_list[f])}
```
2. In most other cases, I favour variable names composed of simple descriptive words, usually not more than two in length, separated by an underscore. Other common separators include a period, or the use of capitalization on the first letter of each word: `like_this`, `or.this`, `OrThis`. Two words is usually a good choice, becuase it offers increased **clarity** as well as reducing the chances that you might select a name which is already being used for some R function.

## Other considerations

If you have a series of similar variables, it's always wise to provide them with similar names. If you're using two-word names, you might use a common element in either the leftmost or rightmost name. The only real significance here is how these items will appear alphabetically in the R Environment tab, where your objects in memory are listed. By putting the common element in the variable names first, these objects will be listed together in the Environment, which may be advantageous (I usually find it helpful).
```{r, eval=F}
female_F1 <- c(...)
male_F1 <- c(...)

Or:

F1_female <- c(...)
F1_male <- c(...)
```
These considerations usually don't seem very important when you start writing your code, as you may only be dealing with a handful of variables and objects, which you can keep and easily differentiate in your own memory quite easily. However, you may find that as you develop a particular piece of code and it grows in size, you can quickly reach a threshold beyond which you find it difficult to easily remember what all your variables refer to, or what their function is within the code.

Another good habit is to annotate your code, and this especially applies to variable names -- *even when they seem totally transparent*.
```{r, eval=F}
# this vector is a list of all the female F1 values
F1_female <- c(...)
```
Thorough annotation is probably the most important thing you can do when it comes to variable names, whether they are named `male_pitch_Hz` or `z`. If your annotations are explicit and detailed, this can go a long way towards discounting the need for both brevity and clarity, even though these are always important naming considerations.

# Common code elements

As I work on projects over time, I find that I tend to reproduce some of the same elements over and over in my R scripts. These can include:

* loading commonly-used packages, e.g. `tidyverse`
* setting your Working Directory

There are a couple solutions to this. The simplest is to put such elements at the very beginning of your scripts, and copy-paste them into new scripts as a sort of template. Alternatively, you could make a new copy of an old script, open it, and just delete everything below the common code elements you want to retain. I've done it both ways, and they involve about the same amount of work.
```{r, eval=F}
library(tidyverse)

setwd("~/UserDocuments/RProjects/ThisProject")
```





## Custom functions

NOTE: add examples from usage in loops

## Custom packages