---
title: "Coding Discussion Group"
author: "Coordinator: Sky Onosson"
date: "409 Tier Building -- Feb. 7, 2020"
output: 
  html_document:
    number_sections: true
---

```{r setup, include = FALSE}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

# Optimizing/Tidying Data

> Preliminaries - load the LIPP vowels data:

```{r lipp data, message=FALSE}
lipp <- read_csv("LIPP.csv")
```


## Drop NA results from data

> Check how many NAs exist by column

```{r check NAs}
# how many NA exist in each column
na_columns <- colSums(is.na(lipp))
na_columns

# extract the list of names for each column
na_column_list <- names(na_columns)
na_column_list
```

> Drop rows with NAs; compare sizes of dataframes

```{r drop NAs}
# select columns from list of names for quantitative data
drop_columns <- c(na_column_list[10:13], na_column_list[19:28])
drop_columns

# remove rows from lipp dataframe if there is NA in "drop_column" list
lipp_drop_na <- drop_na(lipp, drop_columns)
 
# alternative: drop rows with any NAs
lipp_zero_na <- drop_na(lipp)

# create dataframe to compare sizes of original dataframe and after dropping NAs
compare_na <- cbind( # cbind = bind columns together
  # first column: count rows in each dataframe version
  c(nrow(lipp), 
    nrow(lipp_drop_na), 
    nrow(lipp_zero_na)),
  # second column: divide each count by count of original x100, round to 2 decimals
  c(round(nrow(lipp)/nrow(lipp)*100, 2), 
    round(nrow(lipp_drop_na)/nrow(lipp)*100, 2), 
    round(nrow(lipp_zero_na)/nrow(lipp)*100, 2)
    )
  )

# assign column and row names
colnames(compare_na) <- c("Count", "Percent")
rownames(compare_na) <- c("Original data", "Drop selected NAs", "Drop all NAs")

compare_na
```

# Vowel plot types

> Subset out some vowels for clarity, basic point plot

```{r subset vowels}
# select only lax vowels
lipp_lax <- lipp %>% 
  filter (vowel_type == "lax")

# set up table of median F1, F2 values by vowel
medians_lax <- lipp_lax %>%
  group_by(vowel) %>% 
  summarise(F2 = mean(F2), F1 = mean(F1))
```

```{r basic vowel plot}
# basic point plot with reversed x, y axes and points coloured by vowel
ggplot(lipp_lax,
       aes(x = F2, y = F1, color = vowel)) + 
  geom_point() +
  scale_x_reverse() + 
  scale_y_reverse() +
  scale_colour_viridis_d()
```

> Aside: subplot of repeated elements across plots

```{r subplot}
# set up object containing list of common plot elements
subplot <- list(scale_x_reverse(),
                scale_y_reverse(),
                scale_colour_viridis_d(option = "viridis"),
                scale_fill_viridis_c(option = "magma"))
```

> Redo plot with ellipses to reduce impact of outliers on visual display

```{r ellipse plot}
# redo basic plot with ellipses replacing points, labels from median table
ggplot(lipp_lax,
       aes(x = F2, y = F1, color = vowel, label = vowel)) + 
  stat_ellipse() +
  geom_label(data = medians_lax) +
  subplot
```

> Heatmap using 2d density plot

```{r heatmap}
# heatmap using stat_density_2d()
ggplot(lipp,
       aes(x = F2, y = F1, fill = stat(nlevel))) + 
  stat_density_2d(geom = "polygon") +
  subplot + 
  # all vowels retained in data: facet_wrap() 
  facet_wrap(~vowel)
```

# Normalization of vowel formant data

**Normalization procedures rely heavily on Dr. M. Barlaz's workshop: https://marissabarlaz.github.io/portfolio/vowelnormalization/**

> Natural log normalization

```{r ellipses: natural log}
# formants wrapped in log() function
ggplot(lipp_lax,
       aes(x = log(F2), y = log(F1), color = vowel, label = vowel)) + 
  stat_ellipse() +
  geom_label(data = medians_lax) +
  subplot
```

> Log-10 normalization

```{r ellipses: log-10}
# formants wrapped in log10() function
ggplot(lipp_lax,
       aes(x = log10(F2), y = log10(F1), color = vowel, label = vowel)) + 
  stat_ellipse() +
  geom_label(data = medians_lax) +
  subplot
```

> ERB (equivalent rectangular bandwidth) normalization

```{r ellipses: ERB}
# load/install phonR package
require(phonR)

# formants wrapped in normErb() function
ggplot(lipp_lax,
       aes(x = normErb(F2), y = normErb(F1), color = vowel, label = vowel)) + 
  stat_ellipse() +
  geom_label(data = medians_lax) +
  subplot
```

> Bark normalization

```{r ellipses: bark}
# load/install phonR package
require(phonR)

# formants wrapped in normBark() function
ggplot(lipp_lax,
       aes(x = normBark(F2), y = normBark(F1), color = vowel, label = vowel)) + 
  stat_ellipse() +
  geom_label(data = medians_lax) +
  subplot
```

> Mel normalization

```{r ellipses: mel}
# load/install phonR package
require(phonR)

# formants wrapped in normMel() function
ggplot(lipp_lax,
       aes(x = normMel(F2), y = normMel(F1), color = vowel, label = vowel)) + 
  stat_ellipse() +
  geom_label(data = medians_lax) +
  subplot
```

> Z-score/Lobanov normalization

This is a speaker-intrinsic method, so we need to run the function grouped by speaker, and regenerate medians table

```{r ellipses: lobanov}
# formants wrapped in scale() function, group by speaker
z_lipp_lax <- lipp_lax %>% 
         group_by(speaker_id) %>% 
         mutate(F1_z = scale(F1), F2_z = scale(F2))

# set up table of median F1, F2 values by vowel
z_medians_lax <- z_lipp_lax %>%
  group_by(vowel) %>% 
  summarise(F2_z = mean(F2_z), F1_z = mean(F1_z))

# plot using new dataframe & medians table
ggplot(z_lipp_lax,
       aes(x = F2_z, y = F1_z, color = vowel, label = vowel)) + 
  stat_ellipse() +
  geom_label(data = z_medians_lax) +
  subplot
```

> Nearey1

This is a speaker-intrinsic method, so we need to run the function grouped by speaker, and regenerate medians table

```{r ellipses: nearey1}
# load/install phonR package
require(phonR)

# set up new dataframe
n1_lipp_lax <- with(lipp_lax, normNearey1(cbind(F1, F2), group = speaker_id))
colnames(n1_lipp_lax) = c("F1_nearey1", "F2_nearey1")
n1_lipp_lax = cbind(lipp_lax, n1_lipp_lax)

# set up table of median F1, F2 values by vowel
n1_medians_lax <- n1_lipp_lax %>%
  group_by(vowel) %>% 
  summarise(F2_nearey1 = mean(F2_nearey1), F1_nearey1 = mean(F1_nearey1))

# plot using new dataframe & medians table
ggplot(n1_lipp_lax,
       aes(x = F2_nearey1, y = F1_nearey1, color = vowel, label = vowel)) + 
  stat_ellipse() +
  geom_label(data = n1_medians_lax) +
  subplot
```

> Nearey2 and ANAE methods require F0-F3, but F0 not available in LIPP data
